---
title: "Technical Appendix"
author: "Antonio Skilton"
date: "10/12/2016"
output: pdf_document
---

##Cleaning the data
The "flash" dataset contains seven columns and 568 rows. The last two columns are manipulated in the code below so that they are more readable. Each row is a country in a given year.

####Columns:
+ country
+ continent
+ year 
+ life expectancy
+ GDP per capita
+ GDP in billions of dollars
+ population in millions of people


```{r,echo=FALSE, message=FALSE, warning=FALSE, results = "hide"}
libraries <- c("dplyr","tidyr","ggplot2","stringr","reshape2","gridExtra","knitr","stargazer")
lapply(libraries,require,character.only = TRUE)
```

```{r}
flash <- tbl_df(read.csv("4_Flash_Proj_1_Data.csv"))

flash %>%
  mutate(country = str_replace_all(country,"[.,]",""),
         country = as.factor(country),
         gdpPercap = gdpPercap/1000,
         popMill = popThous/1000) %>% 
  select(-popThous) -> flash
```

```{r,echo=FALSE}
flash[1:10,] %>% kable(,caption="First ten rows of manipulated flash data")
```

\newpage

#Life Expectancy

According to the regression results, we see that a life expectancy is highly correlated with GDP per capita. Furthermore, we see that every continent has as higher life expectancy than Africa even when GDP per capita is held constant. Even Oceania, which is represented only by Australia, has a significantly different life expectancy than Africa. 

```{r}
flash %>%
  group_by(country,continent) %>%
  summarise_at(vars(lifeExp:popMill),mean) %>%
  lm(lifeExp ~ poly(gdpPercap,3) + continent,.) -> mod
```

```{r,results='asis',echo=FALSE}
stargazer(mod,
          covariate.labels=c("GDP/capita","(GDP/capita)$^{2}$","(GDP/capita)$^{3}$","Americas",
                             "Asia","Europe","Oceania"),type='latex',no.space=TRUE,header=FALSE)
```

\newpage

Here I create 'flashForPlot', a dataset where the average life expectancy of all five years is taken. 'flashForPlot' is designed to be used by ggplot2. It adds a column named 'fitted', which is the fitted values from the linear regression model above.
```{r,messages=F}
flash %>%
  select(continent,country,lifeExp,gdpPercap,popMill) %>%
  group_by(country,continent) %>%
  summarise_if(is.numeric,mean) %>%
  ungroup %>%
  mutate(predicted = predict(mod),
         difference = lifeExp-predicted,
         label = ifelse(abs(difference) > 9,as.character(country),NA)) -> flashForPlot
```

```{r,echo=FALSE}
flashForPlot[1:10,] %>% kable(,caption="First ten rows of flashForPlot")
```


\newpage

The code below produces the plot below. The plot shows the difference between the fitted from the actual life expectancy value. The fitted values lie on the blue lines. The colored values represent the countries for which their average life expectancy is most different from what would be expected given their GDP per capita and continent.

```{r,fig.align='center'}
flashForPlot %>%
  filter(country != "Australia") %>%
  ggplot() + 
    geom_point(aes(gdpPercap, lifeExp,color=label)) + 
    geom_point(aes(gdpPercap, predicted)) +
    geom_segment(aes(x = gdpPercap, y = lifeExp,
                     xend = gdpPercap, yend = predicted,
                     color = label)) +
    geom_smooth(aes(gdpPercap, predicted),se = FALSE) +
    facet_wrap( ~ continent, scales = "free_x") +
    theme_minimal() + 
    theme(legend.position = "none",
          panel.grid.major.y = element_blank()) +
    xlab("GDP per capita") + 
    ylab("Life expectancy")
```

\newpage

#Convergence
First, we create a vector of the names of the three statistics columns.
```{r}
statNames <- names(select(flash,lifeExp:popMill))
```

```{r,echo=FALSE}
statNames
```


I use the lapply function to calculate growth rates.
```{r}
continentList <- lapply(statNames,function(x){
  flash %>%
    spread_("continent",x) %>%
    group_by(year) %>%
    summarise_at(vars(Africa:Oceania),function(col) median(col,na.rm=T)) %>%
    select(-year) %>% #must remove year because we do not need its growth rate
    as.matrix %>% #growth rate transformation only possible as matrix
    log %>% diff %>% #calculate growth rates as the difference of the log values
    tbl_df %>% #mutate() only possible with tibble (or data frame)
    mutate(year = seq(1977,2007,5)) %>%
    melt("year",1:5,variable.name="continent",value.name=paste(x)) %>%
    tbl_df -> flash2
  if(which(statNames == x) == 1){flash2}#if in first loop, lapply returns three columns
  else{select_(flash2,x)}#if not in first loop, lapply returns newly calculated column
})
```

Using a for loop, I iteratively bind each successive column to the first data frame.
```{r}
for(i in 2:4) continentList[[1]] <- bind_cols(continentList[[1]],continentList[i])
continentGrowthRates <- continentList[[1]]
```

\newpage

```{r}
continentGrowthRates %>%
  filter(continent != "Oceania") %>%
  #gather(stat,growth_rate,lifeExp:popMill) %>%
  ggplot(aes(year,gdpPercap,color=continent)) +
  geom_line() +
  geom_hline(yintercept = 0) +
  #facet_wrap(~continent) + 
  theme_minimal() + 
  xlab("Year") + 
  ylab("Median GDP per capita growth rate")
```

